---
title: "Using DataPackageR"
author: "Greg Finak <gfinak@fredhutch.org>"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    keep_md: TRUE
    toc: yes
vignette: >
  %\VignetteIndexEntry{A guide to using DataPackageR}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  %\usepackage{graphicx}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  eval = TRUE
)
```

## Purpose

This vignette demonstrates how to use DataPackageR to build a data package. DataPackageR aims to simplify data package construction. It provides mechanisms for reproducibly preprocessing and tidying raw data into into documented, versioned, and packaged analysis-ready data sets. Long-running or computationally intensive data processing can be decoupled from the usual `R CMD build` process while maintinaing [data lineage](https://en.wikipedia.org/wiki/Data_lineage).

For demonstration purposes, in this vignette we will subset and package the `mtcars` data set.

## Set up a new data package.

We'll set up a new data package based on the `mtcars` example in the [README](https://github.com/ropensci/DataPackageR/blob/master/README.md).
The `datapackage_skeleton()` API is used to set up a new package. 
The user needs to provide:

- R or Rmd files that process data.
- A list of R object names created by those files.
- Optionally, a path to a directory of raw data (will be copied into the package).
- Optionally, a list of additional files that may be dependencies of your R or Rmd data processing files. 

```{r minimal_example, results='hide', eval = rmarkdown::pandoc_available()}
library(DataPackageR)

# Let's reproducibly package the cars in the mtcars dataset with speed
# > 20. Our dataset will be called `cars_over_20`.

# Get the code file that turns the raw data to our packaged and
# processed analysis-ready dataset.
processing_code <-
    system.file("extdata", "tests", "subsetCars.Rmd", package = "DataPackageR")

# Create the package framework.
DataPackageR::datapackage_skeleton(name = "mtcars20",
                                   force = TRUE,
                                   code_files = processing_code,
                                   r_object_names = "cars_over_20",
                                   path = tempdir() 
                                   #dependencies argument is empty
                                   #raw_data_dir argument is empty.
                                   )

```

### What's in the package skeleton structure?

The process above has created a DataPackageR source tree named "mtcars20" in a temporary directory. For a real use case, you would pick a path on your filesystem where you could then initialize a new github repository for the package.

The contents of `mtcars20` are:

```{r dirstructure,echo=FALSE, eval = rmarkdown::pandoc_available()}
library(data.tree)
df <- data.frame(
  pathString = file.path(
    "mtcars20",
    list.files(
      file.path(tempdir(), "mtcars20"),
      include.dirs = TRUE,
      recursive = TRUE
    )
  )
)
as.Node(df)
```

You should fill out the `DESCRIPTION` file to describe your data package. 
It contains a new `DataVersion` string that will be automatically incremented when the data package is built *if the packaged data has changed*. 

The user-provided code files reside in `data-raw`. They are executed during the data package build process.

### A note about the YAML config file.

A `datapackager.yml` file is used to configure and control the build process.

The contents are:

```{r, echo=FALSE, eval = rmarkdown::pandoc_available()}
cat(yaml::as.yaml(yaml::yaml.load_file(file.path(tempdir(),"mtcars20","datapackager.yml"))))
```

The two main pieces of information in the configuration are a list of the files to be processed and the data sets the package will store.

This example packages an R data set named `cars_over_20` (the name was passed to `datapackage_skeleton()`), which is created by the `subsetCars.Rmd` file. 

The objects must be listed in the yaml configuration file. `datapackage_skeleton()` ensures this is done for you automatically. 

DataPackageR provides an API for modifying this file, so it does not need to be done by hand. 

Further information on the contents of the YAML configuration file, and the API are in the [YAML Configuration Details](YAML_Configuration_Details.html) vignette.

### Where do I put my raw datasets?

Raw data (provided the size is not prohibitive) can be placed in `inst/extdata`.

The `datapackage_skeleton()` API has the `raw_data_dir` argument, which will copy the contents of `raw_data_dir`  (and its subdirectories) into `inst/extdata` automatically. 

In this example we are reading the `mtcars` data set that is already in memory, rather than from the file system.

### An API to read raw data sets from within an R or Rmd procesing script.

As stated in the README, in order for your processing scripts to be portable, you should not use absolute paths to files. DataPackageR provides an API to point to the data package root directory and the `inst/extdata` and `data` subdirectories. These are useful for constructing portable paths in your code to read files from these locations.

For example, to construct a path to a file named "mydata.csv" located in `inst/extdata` in your data package source tree:

- use `DataPackageR::project_extdata_path("mydata.csv")` in your `R` or `Rmd` file. This would return: e.g., `r file.path(tempdir(),"mtcars20","inst","extdata","mydata.csv")`

Similarly: 

- `DataPackageR::project_path()`  constructs a path to the data package root directory. (e.g., `r file.path(tempdir(),"mtcars20")`)
- `DataPackageR::project_data_path()` constructs a path to the data package `data` subdirectory. (e.g., `r file.path(tempdir(),"mtcars20","data")`)

Raw data sets that are stored externally (outside the data package source tree) can be constructed relative to the `project_path()`.

### YAML header metadata for R files and Rmd files.

If your processing scripts are Rmd files, the usual yaml header for rmarkdown documents should be present.

If your processing scripts are R files, you can still include a yaml header, but it should be commented with `#'` and it should be at the top of your R file. For example, a test R file in the DataPackageR package looks as follows:

```
#'---
#'title: Sample report  from R script
#'author: Greg Finak
#'date: August 1, 2018
#'---
data <- runif(100)
```

This will be converted to an Rmd file with a proper yaml header, which will then be turned into a vignette and indexed in the built package.


## Build the data package.

Once the skeleton framework is set up, run the preprocessing code to build `cars_over_20`, and reproducibly enclose it in a package.

```{r , eval = rmarkdown::pandoc_available()}
dir.create(file.path(tempdir(),"lib"))

DataPackageR:::package_build(
  file.path(tempdir(),"mtcars20"),
  install = TRUE,
  lib = file.path(tempdir(),"lib")
)
```

### Documenting your data set changes in NEWS.

When you build a package in interactive mode, you will be prompted to input text describing the changes to your data package (one line). 

These will appear in the NEWS.md file in the following format:

```
DataVersion: xx.yy.zz
========
A description of your changes to the package

[The rest of the file]
```

### Logging the build process.

DataPackageR uses the `futile.logger` package to log progress. 

If there are errors in the processing, the script will notify you via logging to console and to  `/private/tmp/Test/inst/extdata/Logfiles/processing.log`. Errors should be corrected and the build repeated.

If everything goes smoothly, you will have a new package built in the parent directory. 

In this case we have a new package: `mtcars20_1.0.tar.gz`. 

### A note about the package source directory after building.

The package source directory changes after the first build.

```{r, echo=FALSE, eval = rmarkdown::pandoc_available()}
df <- data.frame(
  pathString = file.path(
    "mtcars20",
    list.files(
      file.path(tempdir(), "mtcars20"),
      include.dirs = TRUE,
      recursive = TRUE
    )
  )
)

as.Node(df)
```

### Update the autogenerated documentation. 

After the first build, the `R` directory contains `mtcars.R` that has autogenerated `roxygen2` markup documentation for the data package and for the `cars_over20` packaged data. 

The processed `Rd` files can be found in `man`. 

The autogenerated documentation source is in the `documentation.R` file in `data-raw`. 

You should update this file to properly document your objects. Then rebuild the documentation:

```{r rebuild_docs, eval = rmarkdown::pandoc_available()}
dir.create(file.path(tempdir(),"lib")) # a temporary library directory
document(file.path(tempdir(),"mtcars20"), lib = file.path(tempdir(),"lib"))
```

Updating documentation does not reprocess the data.

Once the the documentation is updated in `R/mtcars.R`, then run `package_build()` again.

### Why not just use R CMD build?

If the processing script is time consuming or the data set is particularly large, then `R CMD build` would run the code each time the package is installed. In such cases, raw data may not be available, or the environment to do the data processing may not be set up for each user of the data. DataPackageR decouples data processing from package building/installation for data consumers.

## Installing and using the new data package.

### Accessing vignettes, data sets, and data set documentation. 

The package source also contains files in the `vignettes` and `inst/doc` directories that provide a log of the data processing. 

When the package is installed, these will be accessible via the `vignette()` API. 

The vignette will detail the processing performed by the `subsetCars.Rmd` processing script. 

The data set documentation will be accessible via `?cars_over_20`, and the data sets via `data()`. 

```{r, eval = rmarkdown::pandoc_available()}
# Create a temporary library to install into.
dir.create(file.path(tempdir(),"lib"))

# Let's use the package we just created.
install.packages(file.path(tempdir(),"mtcars20_1.0.tar.gz"), type = "source", repos = NULL, lib = file.path(tempdir(),"lib"))
lns <- loadNamespace
if (!"package:mtcars20"%in%search())
  attachNamespace(lns('mtcars20',lib.loc = file.path(tempdir(),"lib"))) #use library() in your code
data("cars_over_20") # load the data

cars_over_20 # now we can use it.
?cars_over_20 # See the documentation you wrote in data-raw/documentation.R.
  
vignettes <- vignette(package = "mtcars20", lib.loc = file.path(tempdir(),"lib"))
vignettes$results
```

### Using the DataVersion.

Your downstream data analysis can depend on a specific version of the data in your data package by testing the DataVersion string in the DESCRIPTION file. 

We provide an API for this:

```{r, eval = rmarkdown::pandoc_available()}
# We can easily check the version of the data.
DataPackageR::data_version("mtcars20", lib.loc = file.path(tempdir(),"lib"))

# You can use an assert to check the data version in  reports and
# analyses that use the packaged data.
assert_data_version(data_package_name = "mtcars20",
                    version_string = "0.1.0", acceptable = "equal",
                    lib.loc = file.path(tempdir(),"lib"))  #If this fails, execution stops
                                           #and provides an informative error.
```

# Migrating old data packages.

Version 1.12.0 has moved away from controlling the build process using `datasets.R` and an additional `masterfile` argument. 

The build process is now controlled via a `datapackager.yml` configuration file located in the package root directory. See [YAML Configuration Details](YAML_Configuration_Details.html).

## Create a datapackager.yml file.

You can migrate an old package by constructing such a config file using the `construct_yml_config()` API.

```{r  construct_config, eval = rmarkdown::pandoc_available()}
# Assume I have file1.Rmd and file2.R located in /data-raw, and these
# create 'object1' and 'object2' respectively.

config <- construct_yml_config(code = c("file1.Rmd", "file2.R"),
                               data = c("object1", "object2"))
cat(yaml::as.yaml(config))
```

`config` is a newly constructed yaml configuration object. It can be written to the package directory:

```{r, eval = rmarkdown::pandoc_available()}
path_to_package <- tempdir() # e.g., if tempdir() was the root of our package.
yml_write(config, path = path_to_package)
```

Now the package at `path_to_package` will build with version 1.12.0 or greater.

### Reading data sets from Rmd files.

In versions prior to 1.12.1 we would read data sets from `inst/extdata` in an `Rmd` script using paths relative to `data-raw` in the data package source tree. 

For example:

#### The old way.

```r
# read 'myfile.csv' from inst/extdata relative to data-raw where the Rmd is rendered.
read.csv(file.path("../inst/extdata","myfile.csv"))
```

Now `Rmd` and `R` scripts are processed in `render_root` defined in the yaml config.

To read a raw data set we can get the path to the package source directory using an API call:

#### The new way.

```r
# DataPackageR::project_extdata_path() returns the path to the data package inst/extdata subdirectory directory.
# DataPackageR::project_path() returns the path to the data package root directory.
# DataPackageR::project_data_path() returns the path to the data package data subdirectory directory.
read.csv(DataPackageR::project_extdata_path("myfile.csv"))
```

# Partial builds.

We can also perform partial builds of a subset of files in a package by toggling the `enabled` key in the yaml config file.

This can be done with the following API:

```{r echo=1:2, eval = rmarkdown::pandoc_available()}
config <- yml_disable_compile(config,filenames = "file2.R")
yml_write(config, path = path_to_package) # write modified yml to the package.
cat(yaml::as.yaml(config))
```

Note that the modified configuration needs to be written back to the package source directory in order for the 
changes to take effect. 

The consequence of toggling a file to `enable: no` is that it will be skipped when the package is rebuilt, 
but the data will still be retained in the package, and the documentation will not be altered. 

This is useful in situations where we have multiple data sets, and we want to re-run one script to update a specific data set, but not the other scripts because they may be too time consuming.

# Multi-script pipelines.

We may have situations where we have mutli-script pipelines. There are two ways to share data among scripts. 

1. filesystem artifacts
2. data objects passed to subsequent scripts

## File system artifacts.

The yaml configuration property `render_root` specifies the working directory where scripts will be rendered.

If a script writes files to the working directory, that is where files will appear. These can be read by subsequent scripts.

## Passing data objects to subsequent scripts.

A script can access a data object designated to be packaged by previously ran scripts using `datapackager_object_read()`.

For example, `script2.Rmd` will run after `script1.Rmd`. `script2.Rmd` needs to access a data object that has been designated to be packaged named `dataset1`, which was created by `script1.Rmd`. This data set can be accessed by `script2.Rmd` using the following expression:

`dataset1 <- DataPackageR::datapackager_object_read("dataset1")`. 

Passing of data objects amongst scripts can be turned off via:

`package_build(deps = FALSE)`

# Next steps.

We recommend the following once your package is created.

## Place your package under source control.

You now have a data package source tree. 

- **Place your package under version control**
    1. Call `git init` in the package source root to initialize a new git repository.
    2. [Create a new repository for your data package on github](https://help.github.com/articles/create-a-repo/).
    3. Push your local package repository to `github`. [see step 7](https://help.github.com/articles/adding-an-existing-project-to-github-using-the-command-line/)


This will let you version control your data processing code, and will provide a mechanism for sharing your package with others.

For more details on using git and github with R, there is an excellent guide provided by Jenny Bryan: [Happy Git and GitHub for the useR](https://happygitwithr.com/) and Hadley Wickham's [book on R packages]( https://r-pkgs.org/).

# Additional Details.

## Fingerprints of stored data objects.

DataPackageR calcudevtools::install(build_vignettes = T)lates an md5 checksum of each data object it stores, and keeps track of them in a file
called `DATADIGEST`.

- Each time the package is rebuilt, the md5 sums of the new data objects are compared against `DATADIGEST`.
- If they don't match, the build process checks that the `DataVersion` string has been incremented in the `DESCRIPTION` file.
- If it has not, the build process will exit and produce an error message.

### DATADIGEST

The `DATADIGEST` file contains the following:

```{r, echo=FALSE, eval = rmarkdown::pandoc_available()}
cat(readLines(file.path(tempdir(),"mtcars20","DATADIGEST")),sep="\n")
```

### DESCRIPTION

The description file has the new `DataVersion` string.

```{r echo=FALSE, eval = rmarkdown::pandoc_available()}
cat(readLines(file.path(tempdir(),"mtcars20","DESCRIPTION")),sep="\n")
```




