---
title: "The DataPackageR YAML configuration file."
author: "Greg Finak <gfinak@fredhutch.org>"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    keep_md: TRUE
    toc: yes
  bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{The DataPackageR YAML configuration file.}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  %\usepackage{graphicx}
editor_options: 
  chunk_output_type: inline
---

# Configuring and controlling DataPackageR builds.

Data package builds are controlled using the `datapackager.yml` file. 

This file is created in the package source tree when the user creates a package using `datapackage_skeleton()`. 

It is automatically populated with the names of the `code_files` and `data_objects` the passed in to datapackage_skeleton.

## The `datapackager.yml` file.

The structure of a correctly formatted `datapackager.yml` file is shown below:

```{r, echo = FALSE, results = 'hide', eval = rmarkdown::pandoc_available()}
library(DataPackageR)
library(yaml)
yml <- DataPackageR::construct_yml_config(code = "subsetCars.Rmd", data = "cars_over_20")
```

```{r, echo = FALSE, comment="", eval = rmarkdown::pandoc_available()}
cat(yaml::as.yaml(yml))
```

## YAML config file properties.

The main section of the file is the `configuration:` section.

It has three properties:

- `files:` 
  
  The files (`R` or `Rmd`) to be processed by DataPackageR. They are processed in the order shown. Users running multi-script workflows with dependencies between the scripts need to ensure the files are processed in the correct order. 
  
  Here `subsetCars.Rmd` is the only file to process. The name is transformed to an absolute path within the package.
  
  Each file itself has just one property:
  
  <!-- - `name:`  -->
     <!-- The name of the file. This is transformed to an absolute path within the package. -->
    
  - `enabled:` 
     A logical `yes`, `no` flag indicating whether the file should be rendered during the build, or whether it should be skipped.
    This is useful for 'turning off' long running processing tasks if they have not changed. Disabling processing of a file will not overwrite existing documentation or data objecs created during previous builds.
  
- `objects:` 
  
  The names of the data objects created by the processing files, to be stored in the package. These names are compared against the objects created in the render environment by each file. They names must match.
- `render_root:` 
  
  The directory where the `Rmd` or `R` files will be rendered. Defaults to a randomly named subdirectory of `tempdir()`. Allows workflows that use multiple scripts and create file system artifacts to function correctly by simply writing to and reading from the working directory.

## Editing the YAML config file.

The structure of the YAML is simple enough to understand but complex enough that it can be a pain to edit by hand.

DataPackageR provides a number of API calls to construct, read, modify, and write the yaml config file.

### API calls

#### `construct_yml_config` 

  Make an r object representing a YAML config file.
  
##### Example
  The YAML config shown above was created by: 
```{r, eval = rmarkdown::pandoc_available()}
# Note this is done by the datapackage_skeleton. 
# The user doesn't usually need to call 
# construct_yml_config()
yml <- DataPackageR::construct_yml_config(
  code = "subsetCars.Rmd",
  data = "cars_over_20"
  )
```
  
  
#### `yml_find` 

  Read a yaml config file from a package path into an r object.
  
##### Example
  Read the YAML config file from the `mtcars20` example.
  
```{r eval=FALSE}
# returns an r object representation of
# the config file.
mtcars20_config <- yml_find(
  file.path(tempdir(),"mtcars20")
  )
```

#### `yml_list_objects` 

  List the `objects` in a config read by `yml_find`.
  

##### Example

```{r, comment="", eval = rmarkdown::pandoc_available()}
  yml_list_objects(yml)
```
  
#### `yml_list_files` 

  List the `files` in a config read by `yml_find`.

##### Example

```{r, comment="", eval = rmarkdown::pandoc_available()}
  yml_list_files(yml)
```
  
#### `yml_disable_compile` 

  Disable compilation of named files in a config read by `yml_find`.

##### Example

```{r, comment="", echo = 1, eval = rmarkdown::pandoc_available()}
yml_disabled <- yml_disable_compile(
    yml,
    filenames = "subsetCars.Rmd")
cat(as.yaml(yml_disabled))
```

#### `yml_enable_compile` 

  Enable compilation of named files in a config read by `yml_find`.

##### Example

```{r, comment="", echo = 1, eval = rmarkdown::pandoc_available()}
yml_enabled <- yml_enable_compile(
    yml,
    filenames = "subsetCars.Rmd")
cat(as.yaml(yml_enabled))
```

#### `yml_add_files` 
  
  Add named files to a config read by `yml_find`.
  
##### Example

```{r, comment="", echo = 1, eval = rmarkdown::pandoc_available()}
yml_twofiles <- yml_add_files(
    yml,
    filenames = "anotherFile.Rmd")
# cat(as.yaml(yml_twofiles))
```

#### `yml_add_objects` 

  Add named objects to a config read by `yml_find`.

##### Example

```{r, comment="", echo = 1, eval = rmarkdown::pandoc_available()}
yml_twoobj <- yml_add_objects(
    yml_twofiles,
    objects = "another_object")
# cat(as.yaml(yml_twoobj))
```

#### `yml_remove_files` 

  Remove named files from a config read by `yml_find`.

##### Example

```{r, comment="", echo = 1, eval = rmarkdown::pandoc_available()}
yml_twoobj <- yml_remove_files(
    yml_twoobj,
    filenames = "anotherFile.Rmd")
# cat(as.yaml(yml_twoobj))
```

#### `yml_remove_objects` 

  Remove named objects from a config read by `yml_find`.

##### Example

```{r, comment="", echo = 1, eval = rmarkdown::pandoc_available()}
yml_oneobj <- yml_remove_objects(
    yml_twoobj,
    objects = "another_object")
# cat(as.yaml(yml_oneobj))
```

#### `yml_write` 

  Write a modified config to its package path.

##### Example

```{r, eval = FALSE}
yml_write(yml_oneobj, path = "path_to_package")
```

The `yml_oneobj` read by `yml_find()` carries an attribute
that is the path to the package. The user doesn't need to pass a `path` to `yml_write` if the config has been read by `yml_find`.
